### 2012-11-26 branches/CUDA/BookGPU/Chapters/chapter14/ch14.tex
updates to gpu chapter
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1820 40876441-3496-0410-a613-a2fca35b2ae2

\chapter{Ludwig}

@@

\section{Introduction}
The lattice Boltzmann (LB) method (for an overview see, e.g.,
\cite{succi-book}) has become a popular approach to a variety of fluid
dynamics problems.  It provides a way to solve the incompressible,
isothermal Navier-Stokes equations and has the attractive features of
being both explicit in time and local in space. This makes the LB
method well-suited to parallel computation. Many efficient parallel
implementations of the LB method have been undertaken, typically using
a combination of distributed domain decomposition and the Message
Passing Interface MPI \cite{mpi-standard}. However, the potential
performance benefits offered by GPUs has motivated a new `mixed-mode'
approach to address verly large problems. Here, fine-grained
parallelism is implemented on the GPU, while MPI is reserved for
larger-scale parallelism.  This mixed mode is of increasing interest
to application programmers as many supercomputing services are moving
toward clusters of GPU accelerated nodes.

The \textit{Ludwig} code \cite{desplat} is a LB package developed
specifically for complex fluid problems, for example mixtures,
surfactants, liquid crystals, and particle
suspensions~\cite{aidun2010}.  In this chapter we present the steps
required to allow Ludwig to efficiently exploit many NVIDIA GPUs in
parallel.  We show that Ludwig scales excellently to at least the
thousand GPU level (the largest resource available at the time of
writing) with indications that the code will scale to much larger
systems as they come on-line. We first consider the problem of a
binary fluid mixture (e.g., oil and water) modelled using the LB
approach, and we go on to describe how we deal with the complexities
arising when colloidal particles are introduced to interact with the
fluid.

In Section \ref{ch14:sec:background} we give background details to
introduce the the methods being employed. In Section
\ref{ch14:sec:singlegpu}, we describe the adaptations made to Ludwig
to allow it to efficiently exploit the GPU architecture. In Section
\ref{ch14:sec:singlegpu}, we extend this description to the work
required to allow exploitation of many GPUs in parallel. We present
performance and scaling results, for the binary fluid case, in Section
\ref{ch14:sec:performance}, before going, in Section
\ref{ch14:sec:particles}, to describe how we minimise the overheads
involved with introducing colloidal particles.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{Chapters/chapter14/figures/basiclattice}
\caption{The lattice Boltzmann approach discritises space into a 3D lattice (left). The fluid is represented at each lattice site (right), by the {\it distribution} vector: each component corresponding to a specific velocity direction. Here we are interested in the {\it D3Q19} model, in which there are 3 spatial directions and 19 velocity components (including a zero component).}
\label{ch14:fig:basiclattice}
\end{figure}

A symmetric binary fluid, in which both components have the same
density and viscosity, may be described by introducing a composition
variable or \textit{order parameter} $\phi(\mathbf{r})$. This is
a function of position $\mathbf{r}$ and describes the relative
proportions of the two fluids present locally. To describe the
thermodynamics of the system it is possible to write down a
\textit{free energy} which is a functional of the order parameter.
The equation of motion for $\phi$ is then
\begin{equation}
\partial_t \phi + \mathbf{\nabla}.(\mathbf{u}\phi) = 
- \mathbf{\nabla} . (M \mathbf{\nabla}\mu),
\label{cahn-hilliard}
\end{equation}
where $\mu$ is the chemical potential related to the functional derivative
of the free energy, and $M$ is a mobility. This couples
to the fluid velocity field $\mathbf{u}(\mathbf{r})$, whose evolution
obeys the Navier-Stokes equations describing the conservation of mass
(or density $\rho$) and momentum
\begin{equation}
\rho  [ \partial_t \mathbf{u} + (\mathbf{u}.\mathbf{\nabla})\mathbf{u} ]
= -\nabla p + \eta \nabla^2 \mathbf{u} + \mathbf{f}(\mathbf{r}),
\end{equation}
where $p$ is the isotropic pressure and $\eta$ is the viscosity.
A force density
$\mathbf{f}(\mathbf{r})$  describes the force exerted by a
curved fluid interface on the fluid. It is sufficient to note
here that this force depends locally on the order parameter and its
derivatives $\nabla\phi$ and $\nabla^2 \phi$. For a full
description the interested reader is referred to, e.g.,
\cite{bray1994,kendon2001}.

The LB approach makes use of a regular 3-dimensional
lattice (see Figure \ref{ch14:fig:basiclattice}) with discrete spacing $\Delta r$. It also makes use of a
discrete velocity space $\mathbf{c}_i$, where the $\mathbf{c}_i$
are chosen to capture the correct symmetries of the Navier-Stokes
equations. A typical choice, used here, is the so-called D3Q19
basis in three dimensions where there is one velocity such that
$\mathbf{c} \Delta t$ is zero, along with six extending to the nearest
neighbour
lattice sites, and twelve extending to the next-nearest neighbour sites
($\Delta t$ being the discrete time step). The fundamental object
in LB is then the distribution function $f_i (\mathbf{r};t)$ whose
moments are related to the local hydrodynamic quantities: the fluid
density, momentum, and stress. The time evolution of the distribution
function is described by a discrete Boltzmann equation
\begin{equation}
f_i(\mathbf{r} + \mathbf{c}_i \Delta t; t) - f_i(\mathbf{r}; t) 
= - {\cal L}_{ij} f_j(\mathbf{r};t).
\end{equation}
It is convenient to think of this in two stages. First, the right hand
side represents the action of a collision operator ${\cal L}_{ij}$,
which is local to each lattice site and relaxes the distribution toward
a local equilibrium at a rate ultimately related to the fluid viscosity.
Second, the left hand side represents a propagation step (sometimes referred
to as streaming step), in which each element $i$ of the distribution is
displaced $\mathbf{c}_i \Delta t$, i.e., one lattice spacing in the
appropriate direction per discrete time step.

Computationally, we store a vector of 19 double precision floating
point values at each lattice site for the distribution function $f_i$.
The collision operation, which is local at each lattice site, may be
thought of as follows. The matrix-vector multiplication $L_{ji}f_i$ is
used to transform the distributions into the hydrodynamic quantities,
where $L_{ij}$ is a constant 19x19 matrix related to the choice of
$\mathbf{c}_i$. The non-conserved hydrodynamic quantities are then
relaxed toward their (known) equilibrium values, and are transformed
back to new post-collision distributions via the inverse transformation
$L^{-1}_{ij}$. This gives rise to the need for a minimum of 2x19$^2$
floating point multiplications per lattice site. In contrast, the
propagation stage consists solely of moving the distribution values
between lattice sites, and involves no floating point operations.

To represent the composition variable, we introduce a second distribution
function $g_i (\mathbf{r};t)$ whose first moment is $\phi$, second moment
is the order parameter flux, and so on \cite{swift1996}.
The time evolution of this second distribution follows an analogous
collision and propagation process to that described for $f_i$, and
so provides a solution to Eq.~\ref{cahn-hilliard} \cite{stratford-jsp2005}.

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{Chapters/chapter14/figures/decomphalo}
\caption{Left: the lattice is decomposed between MPI tasks. For clarity we show a 2D decomposition of a 3D lattice, but in practice we decompose in all 3 dimensions. Halo cells are added to each sudomain, as shown on the right for a single slice, which store data retrieved from remote neighbours in the halo exchange.}
\label{ch14:fig:decomphalo}
\end{figure}

As illustrated in Figure \ref{ch14:fig:decomphalo}, Ludwig is
parallelized (across CPU cores) using a regular 3D domain decomposition
where communications are performed via message passing using MPI.  We
note that while the collision is local for $f_i$, computation of the
force terms requires the evaluation of the derivatives of $\phi$ via a
standard finite different stencil.  This gives rise to the need for a
halo exchange involving the scalar $\phi$ before the collision.
Before the propagation stage, a halo exchange in both $f_i$ and $g_i$
is required.

For a 3D halo exchange, 2D planes of data are exchanged in each of the
positive and negative directions, for each of the 3 dimensions, i.e. 6
planes are sent and received by each process.  For the CPU version of
Ludwig, MPI datatype functionality is used to designate these planes.
The $\phi$ halo exchange involves scalar values at each site, whereas
the distribution halo exchange involves a higher volume of data since
we must transfer multiple components of each velocity vector. However
the number of velocity components included in a given transfer can be
filtered from 19 to 5 (for D3Q19), since only those elements
propagating outward from a local domain play a role in the propagation
stage. The specific components required depend on the direction of
transfer, and again MPI datatype functionality is used to designate
this.

\section{GPU Implementation}\label{ch14:sec:singlegpu}

In this section we describe the methods used to enable Ludwig
for the GPU architecture and describe some key optimizations that
significantly enhance performance.

To ensure that \textit{Ludwig} remains suitable for CPU-based systems,
we augment rather than modify the original code. All new GPU-related
functionality are implemented in an additive manner, mirroring the
modular structure of the original code. The new GPU modules have
interfaces similar to the original CPU functions, and the ability to
switch between versions is achieved at compile-time via C
pre-processing directives.  All GPU compute kernels are implemented using
CUDA.  Wrapper routines are developed to specify the
decomposition, invoke the CUDA kernels and perform the necessary data
management operations: the latter are facilitated through newly
developed modules for creation and destruction of device memory
structures, the copying of data between host and device, and where
necessary the packing and buffering of data.

The most important code section, in terms of performance, is the
collision stage. But it is important to offload all other
computational components of the timestep. Kernels with relatively low
computational demand, such as the propagation stage and operations
involving the order parameter, are also GPU enabled: this allows data
to remain resident on the GPU, and avoids expensive host-device data
transfer at each timestep (which would reduce the benefit of GPU
acceleration).  In order to port to the GPU in an incremental fashion,
while retaining the ability to test for correctness throughout the
process, we require a suite of data management functions. For each
data structure (such as the distribution, which we will use as an
example), we maintain a copy in both the CPU and GPU memory spaces.
Functions such as \verb+put_dist_on_gpu()+, and
\verb+get_dist_from_gpu()+, which abstract underlying CUDA
functionality, allow us to update the GPU copy from the host copy and
vice versa\footnote {Within the implementation of these functions we
  have to work around data encapsulation issues. In the code, the use
  of the file scope identifier \texttt{static} prevents direct access
  to data; instead access functions are provided.  For the GPU code,
  it was necessary to use these existing access functions to copy
  these data to equivalent arrays on the device (via temporary buffers
  on the host).  Within the CUDA kernel, the data access function
  calls are substituted with array accesses.  The process is reversed
  when encapsulated data needs to be transferred back to the host.}.
This makes it easy, in the function implementing the Ludwig timestep,
to switch between the CPU and GPU for different components in the
timestep: we can develop GPU functionality incrementally while
retaining a code that runs correctly (albeit slowly due to data
transfer overheads).  Once every component in the timestep is
GPU-enabled, then it becomes possible to move such data transfers
outside the timestep loop, and just use the GPU-resident versions
throughout the simulation.

To achieve optimal performance, it is vital to fully exploit
parallelism inherent to the NVIDIA architecture, particularly for
those matrix-vector operations within the dominant collision kernel. A
simplified schematic of a serial implementation of such a matrix-vector
operations is as follows:
{\footnotesize
\begin{verbatim}
// loop over lattice sites
for ( is = 0; is < N; is++){
  ...
  //load distribution from site-major structure
  for (i = 0; i < 19; i++)    
    f[i]=f_all[is][i]
  ...
  //perform matrix-vector multiplication  
  for (i = 0; i < 19; i++){    
    a[i] = 0.0;    
    for (j = 0; j < 19; j++){      
      a[i] += f[j]*L[i][j];   
    }
  }
  ...
}
\end{verbatim}
}
The GPU architecture features a hierarchy of parallelism. At the
lowest level, groups of 32 threads called {\it warps} operate in
lock-step on different data elements - this is SIMD style vector-level
parallelism. Multiple warps are combined into a thread block (in which
communication and synchronisation are possible), and multiple blocks can
run concurrently across the SMs in the GPU (with no communication or
synchronisation possible across blocks).  To decompose on the GPU, we
must choose which of the above loops to assign to which level of
parallelism.

There exists parallelism within the matrix-vector operation, but the
length of each {\it distribution} vector, at 19, is smaller than the
warp size and typical thread block sizes. So therefore we simply
decompose the loop over lattice sites to all levels of parallelism,
i.e.  we use a separate CUDA thread for every lattice site, and each
thread performs the full matrix-vector operation. We find that a block
size of 256 performs best: we decompose our lattice domain into groups
of 256 sites, and assign each group to a block of CUDA threads.

Since the matrix \verb+L+ is constant across CUDA threads (sites), it
can reside in the fast {\it constant} on-chip device memory. For each
\verb+is+, \verb+L+ is multiplied by the distribution function at that
site; the distributions for the complete lattice are stored in
\verb+f_all+ residing in off-chip device global memory.

Two distinct optimizations were found to improve the performance of this code section, and hence the application as a whole, on
the GPU.  First, the original code uses the output vector \verb+a+ to
hold each temporary summation in the inner loop.  The code was adapted
so that a temporary scalar variable was used instead. This allows
caching of intermediate values on-chip and thus avoids expensive
global memory accesses within the inner loop.  Secondly, it can be
seen the original data layout of \verb+f_all+ is site-major, i.e. for
each site, consecutive velocity components are contiguous in memory.
This layout is optimal when executing on the CPU, but is sub-optimal
on the GPU when combined with our CUDA decomposition. An architectural
constraint of NVIDIA GPUs means that optimal global memory bandwidth
is only achieved when data is structured such that threads within a
{\it half-warp} (a group of 16 threads) load data from the same memory
segment in a single transaction: this is referred to as {\it memory
  coalescing}.  Otherwise, memory accesses are serialized,
significantly degrading performance. The coalescing criteria is met
when consecutive threads read consecutive memory addresses. Therefore
the alternative transposed velocity-major layout of \verb+f_all+ is
more optimal, i.e.  for each velocity component, consecutive site
values are contiguous in memory.  \textit{Ludwig} was modified to
allow a choice of distribution data layout at compilation time: the
original (most optimal for the CPU) or that permitting coalesced
accesses on the GPU.

A schematic of our optimised CUDA code, to be compared with the
listing above, is
{\footnotesize
\begin{verbatim}
// calculate is from CUDA thread and block indexes
...
//load distribution from velocity-major structure
for (i = 0; i < 19; i++)    
   f[i]=f_all[i][is]
...
//perform matrix-vector multiplication  
for (i = 0; i < 19; i++){    
 double a_tmp = 0.0;    
 for (j = 0; j < 19; j++){      
   a_tmp += f[j]*L[i][j];   
 }
 a[i] = a_tmp;    
}
...
\end{verbatim}
}

\section{Multi-GPU Implementation}\label{ch14:sec:parallelgpu}

To execute on multiple GPUs in parallel we decompose the problem using
the existing high-level parallelisaton framework: each MPI task acts
as a host to a separate GPU, with each GPU kernel operating on the
subset of the lattice volume corresponding to the partition assigned
to that task.  As described in Section \ref{ch14:sec:background}, halo
data exchanges are required at each timestep.  Significant
developments are made for the GPU adaptation of this communication
phase to achieve good inter-GPU communication performance and allow
scaling to many GPUs.

The
traditional code uses MPI datatype functionality to designate the six
planes of lattice data to be sent/received, and performs the
communications {\it in-place}, i.e.  there is no explicit buffering of
data in the application.  For the GPU adaptation, the halo planes must
be exchanged between GPUs, on which MPI functionality is not
available, so transfers must be staged through the host CPUs. Buffer
packing/unpacking of the halo planes is explicitly performed using
CUDA kernels, and CUDA memory copy calls are used to transfer data
from a device to its host (using pinned host memory) through the PCI-e
bus (and vice versa). MPI calls are used to transfer the buffers
between hosts. The CPU version of Ludwig has a mechanism, using MPI built-in
data types, to reduce the amount of data communicated by around a
factor of four by only sending those elements of the distribution
propagating outward from a local domain. For the GPU adaptation, this
filtering of data is explicitly performed (again since MPI
functionality is not available on the GPU) in the buffer
packing/unpacking kernels.  Special care had to be taken for those
lattice sites on the corners of the halo planes, which had to be sent
in more than one direction (since the propagation includes accesses to
diagonal neighbours).

The fact that we are transferring multiple buffers allows us to
overlap the communication stages, which each stress different parts of
the hardware, and hence reduce overall communication overhead.  For
example, after the planes corresponding to the first dimension have
been retrieved by the host, these can be exchanged using MPI between
hosts at the same time as the packing and retrieving of the second
dimension planes. The operations are not completely independent across
dimensions because of updates to corner data (see above). We use a
separate CUDA stream for each dimension and overlap to the extent
allowed. When overlapping is invoked, we observe that some of the
host-device communication time can be effectively ``hidden'' behind
the host-host MPI communication time, resulting in an overall speedup.
The effect is more pronounced for the smaller local lattice size,
perhaps due to less CPU memory bandwidth contention. Hence the
optimization is particularly valuable to aid strong scaling, where the
local system size reduces as the number of GPUs increases.

\section{Performance}\label{ch14:sec:performance}

>>>>

\chapter{Ludwig: GPU implementation of a complex fluid lattice Boltzmann
application}

@@

\section{Introduction}
The lattice Boltzmann (LB) method (for an overview see, e.g.,
\cite{succi-book}) has become a popular approach to a variety of fluid
dynamics problems.  It provides a way to solve the incompressible,
isothermal Navier-Stokes equations and has the attractive features of
being both explicit in time and local in space. This makes the LB
method well-suited to parallel computation. Many efficient parallel
implementations of the LB method have been undertaken, typically using
a combination of distributed domain decomposition and the Message
Passing Interface (MPI). However, the potential
performance benefits offered by GPUs has motivated a new `mixed-mode'
approach to address very large problems. Here, fine-grained
parallelism is implemented on the GPU, while MPI is reserved for
larger-scale parallelism.  This mixed mode is of increasing interest
to application programmers at a time when many supercomputing services
are moving
toward clusters of GPU accelerated nodes. The design questions which
arise when developing a lattice Boltzmann code for this type of
heterogeneous system are therefore worth studying. However, similar
questions also recur in many other types of stencil-based algorithms.

The first applications of LB on GPUs were to achieve fluid-like
effects in computer animation, rather than scientific applications per
se.  These early works include simple fluids~\cite{wei2004}, miscible
two-component flow~\cite{zhu2006}, and various image processing tasks
based on the use of partial differential equations~\cite{zhao2007}.
While these early works used relatively low level graphics APIs, the
first CUDA runtime interface implementation was a two-dimensional
simple fluid problem~\cite{toelke2010}.  Following pioneering work on
clusters of GPUs coupled via MPI to study air
pollution~\cite{fan2004}, more recent work has included mixed OpenMP
and CUDA~\cite{myre2011}, Posix threads and CUDA~\cite{obrecht2011},
and MPI and CUDA for increasingly large GPU clusters
\cite{bernaschi2010,xian2011,feichtinger2011}. The heterogeneous
nature of these systems has also spurred interest in approaches
including automatic code generation \cite{walshsaar2012} and autotuning
\cite{williams2011} to aid application performance.

Many of these authors make use of
another attractive feature of LB: the ability to include fixed
solid-fluid boundary conditions as a straightforward addition to
the algorithm to study, for example, flow in
porous media. This points to an important
application area for LB methods: that of complex fluids.
Complex fluids include mixtures, surfactants, liquid crystals,
and particle suspensions, and typically require additional physics
beyond the bare Navier-Stokes equations to provide a full
description~\cite{aidun2010}. The representation of this extra
physics raises additional design questions for the application
programmer. Here, we consider the \textit{Ludwig} code \cite{desplat},
an LB application developed specifically for complex fluids
(\textit{Ludwig} was named for Boltzmann, 1844--1906).
We will present the steps
required to allow Ludwig to exploit efficiently both a single
GPU, and also many GPUs in parallel.  We show that
Ludwig scales excellently to at least the one thousand GPU level
(the largest resource available at the time of writing) with
indications that the code will scale to much larger systems as
they become available. In addition, we consider the steps required
to represent fully-resolved moving solid particles (usually referred
to as colloids in the context of complex fluids). Such particles
need to have their surface resolved on the lattice scale in order
to include relevant surface physics, and must be able to move: e.g.,
to execute Brownian motion in response to random forces from the
fluid. Standard methods are available to represent such particles
(e.g., \cite{ladd1994,nguyen2002}) which are amenable to effective
domain decomposition and message passing \cite{stratford2008}.
We describe below the additional considerations which arise
for such moving particles when developing an
implementation on a GPU.

In the following section we provide a brief overview of the lattice
Boltzmann method, and mention some of the general issues which can
influence performance in the CPU context.
In Section \ref{ch14:sec:singlegpu}, we then describe the alterations
which are required to exploit the GPU architecture effectively, and
highlight how the resulting code differs from the CPU version. In
Section \ref{ch14:sec:parallelgpu}, we extend this description to
include the steps required to allow exploitation of many GPUs in
parallel while retaining effective scaling. We also present results for
a typical benchmark for a fluid-only problem in Section
\ref{ch14:sec:parallelgpu} to demonstrate the success of the approach.
In Section \ref{ch14:sec:particles}, we describe the design choices
which are relevant to a GPU implementation of moving particles. Finally,
we include in the summary a number of general observations on software
engineering and maintenance which arise from our experience.

@@

%\begin{figure}[!t]
%\centering
%\includegraphics[width=12cm]{Chapters/chapter14/figures/basiclattice}
%\caption{The lattice Boltzmann approach discritises space into a 3D lattice (left). The fluid is represented at each lattice site (right), by the {\it distribution} vector: each component corresponding to a specific velocity direction. Here we are interested in the {\it D3Q19} model, in which there are 3 spatial directions and 19 velocity components (including a zero component).}
%\label{ch14:fig:basiclattice}
%\end{figure}

For a general complex fluid problem the starting point is the
fluid velocity field $\mathbf{u}(\mathbf{r})$, whose evolution
obeys the Navier-Stokes equations describing the conservation of mass
(or density $\rho$), and momentum:
\begin{equation}
\rho  [ \partial_t \mathbf{u} + (\mathbf{u}.\mathbf{\nabla})\mathbf{u} ]
= -\nabla p + \eta \nabla^2 \mathbf{u} + \mathbf{f}(\mathbf{r}),
\end{equation}
where $p$ is the isotropic pressure and $\eta$ is the viscosity.
A local force $\mathbf{f}(\mathbf{r})$ provides a means for coupling
to other complex fluid constituents, e.g., it might represent the force
exerted on the fluid by a curved interface between different phases or
components.

The LB approach makes use of a regular three-dimensional
lattice (see Figure \ref{ch14:fig:decomphalo}) with discrete spacing
$\Delta r$. It also makes use of a
discrete velocity space $\mathbf{c}_i$, where the $\mathbf{c}_i$
are chosen to capture the correct symmetries of the Navier-Stokes
equations. A typical choice, used here, is the so-called D3Q19
basis in three dimensions where there is one velocity such that
$\mathbf{c} \Delta t$ is zero, along with six extending to the nearest
neighbour
lattice sites, and twelve extending to the next-nearest neighbour sites
($\Delta t$ being the discrete time step). The fundamental object
in LB is then the distribution function $f_i (\mathbf{r};t)$ whose
moments are related to the local hydrodynamic quantities: the fluid
density, momentum, and stress. The time evolution of the distribution
function is described by a discrete Boltzmann equation
\begin{equation}
f_i(\mathbf{r} + \mathbf{c}_i \Delta t; t) - f_i(\mathbf{r}; t) 
= - {\cal L}_{ij} f_j(\mathbf{r};t).
\end{equation}
It is convenient to think of this in two stages. First, the right hand
side represents the action of a collision operator ${\cal L}_{ij}$,
which is local to each lattice site and relaxes the distribution toward
a local equilibrium at a rate ultimately related to the fluid viscosity.
Second, the left hand side represents a propagation step (sometimes referred
to as streaming step), in which each element $i$ of the distribution is
displaced $\mathbf{c}_i \Delta t$, i.e., one lattice spacing in the
appropriate direction per discrete time step.

More specifically, we store a vector of 19 double precision floating
point values at each lattice site for the distribution function
$f_i(\mathbf{r};t)$.
The collision operation ${\cal L}_{ij}$, which is local at each lattice
site, may be thought of as follows. A matrix-vector multiplication
${\cal M}_{ij}f_j$ is used to transform the distributions into the
hydrodynamic quantities, where ${\cal M}_{ij}$ is a constant 19x19
matrix related to the choice of
$\mathbf{c}_i$. The non-conserved hydrodynamic quantities are then
relaxed toward their (known) equilibrium values, and are transformed
back to new post-collision distributions via the inverse transformation
${\cal M}^{-1}_{ij}$. This gives rise to the need for a minimum of 2x19$^2$
floating point multiplications per lattice site. In contrast, the
propagation stage consists solely of moving the distribution values
between lattice sites, and involves no floating point operations.

In the CPU version, the \textit{Ludwig} implementation stores one
time level of distribution values $f_i(\mathbf{r}; t)$. This distribution
is stored as a flat array of C data type \texttt{double}, and laid out so
that all elements of the velocity are contiguous at a given site (often
referred to as ``array-of-structures'' order). This is appropriate
for sums over
the distribution required locally at the collision stage as illustrated
schematically in Listing~\ref{ch14:listing1}. (The temporary scalar
\texttt{a\_tmp} allows caching of the intermediate accumulated value
in the innermost loop.)
A variety of standard sequential optimisations are relevant for the
collision stage (loop unrolling, inclusion of SIMD operations, and
so on \cite{wellein2006}). 
For example, the collision stage in \textit{Ludwig} includes explicit
SIMD code, which is useful if the compiler on a given platform cannot
identify it.
The propagation stage is separate, and is organised as a ``pull''
(again, various optimisation have been considered, e.g., 
\cite{pohl2003,mattila2007,wittmann2012}).
No further optimisation is done
here beyond ensuring that the ordering of the discrete velocities allows
memory access to be as efficient as possible. While these optimisations
are important, it should be remembered that for some complex fluid
problems, the hydrodynamics embodied in the LB calculation is a relatively
small part of the total computational cost (at the 10\%-level in some cases).
This means optimisation effort may be better concentrated elsewhere.

\begin{lstlisting}[float, label=ch14:listing1,
caption = Collision schematic for CPU.]
/* loop over lattice sites */
for (is = 0; is < nsites; is++) {
  ...
  /* load distribution from ``array-of-structures'' */
  for (i = 0; i < 19; i++)    
    f[i] = f_aos[19*is + i]
  ...
  /* perform matrix-vector multiplication */  
  for (i = 0; i < 19; i++) {    
    a_tmp = 0.0;    
    for (j = 0; j < 19; j++) {      
      a_tmp += f[j]*M[i][j];   
    }
    a[i] = a_tmp;
  }
  ...
}
\end{lstlisting}

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{Chapters/chapter14/figures/decomphalo}
\caption{Left: the lattice is decomposed between MPI tasks. For clarity
we show a 2D decomposition of a 3D lattice, but in practice we decompose
in all 3 dimensions. Halo cells are added to each sub-domain (as shown on
the right for a single slice) which store data retrieved from remote
neighbours in the halo exchange. CAN WE ADD: Lower right: the D3Q19
velocity set required 5 ``outgoing'' elements to be transferred at
each edge.}
\label{ch14:fig:decomphalo}
\end{figure}

The regular 3D decomposition is illustrated in Fig.~\ref{ch14:fig:decomphalo}.
Each local sub-domain is surrounded by a halo, or ghost, region of width one
lattice site. While the collision is local, elements of the distribution
must be exchanged at the edges of the domains to facilitate the propagation.
To achieve the full 3D halo exchange, the standard approach of shifting the
relevant data in each co-ordinate direction in turn is adopted. This
requires appropriate synchronisation, i.e., a receive in the the first
co-ordinate direction must be complete before a send in the second direction
involving relevant data can take place, and so on. We note that only
``outgoing'' elements of the distribution need to be sent at each edge.
For the D3Q19 model, this reduces the volume of data traffic from 19 to
5 of the $f_i(\mathbf{r};t)$ per lattice site at each edge. In the CPU
version, the necessary transfers are implemented in place using
a vector of appropriately strided MPI datatypes for each direction.

\section{Single GPU Implementation}\label{ch14:sec:singlegpu}

In this section we describe the steps taken to enable \textit{Ludwig}
for the GPU. There are a number of crucial issues: first, the
minimisation of data traffic between host and device; second, the
maximisation of available parallelism and third, the issue of
memory coalescing. We discuss each of these in turn.

While the most important section of the code in terms of floating-point
performance is the collision stage, this cannot be the only consideration
for a GPU implementation. It is essential to offload all computational
activity which involves the main data structures (such as the distribution)
to the GPU. This includes kernels with relatively low computational demand,
such as the propagation stage. All relevant data then remain resident on
the GPU, which avoids expensive host-device data transfer at each iteration
of the algorithm. Such transfers would negate any benefit of GPU acceleration.
We note that for a complex fluid code, this requirement can extend to a
considerable number of kernels, although we limit the discussion to
collision and propagation for brevity.

To achieve optimal performance, it is vital to exploit fully the
parallelism inherent in the GPU architecture, particularly for
those matrix-vector operations within the collision kernel.
The GPU architecture features a hierarchy of parallelism. At the
lowest level, groups of 32 threads (warps) operate in
lock-step on different data elements: this is SIMD style vector-level
parallelism. Multiple warps are combined into a thread block (in which
communication and synchronisation are possible), and multiple blocks can
run concurrently across the streaming multiprocessors in the GPU
(with no communication or
synchronisation possible across blocks).  To decompose on the GPU, we
must choose which part of the collision to assign to which level of
parallelism.

While there exists parallelism within the matrix-vector operation, the
length of each vector (here, 19) is smaller than the warp size and typical
thread block sizes. So we simply decompose the loop over lattice sites
to all levels of parallelism, i.e., we use a separate CUDA thread for
every lattice site, and each thread performs the full matrix-vector
operation. We find that a block size of 256 performs well on current
devices: we therefore decompose
the lattice into groups of 256 sites, and assign each group to a block
of CUDA threads. As the matrix ${\cal M}_{ij}$ is constant, it is
assigned to the fast {\it constant} on-chip device memory.

For the propagation stage, the GPU implementation adds a second time
level of distribution values. The data-dependencies inherent in the
propagation mean that the in-place propagation of the CPU version
cannot be parallelised effectively without the additional time level.
As both time levels may remain resident on the GPU, this is not a
significant overhead.

An architectural
constraint of GPUs means that optimal global memory bandwidth
is only achieved when data are structured such that threads within a
{\it half-warp} (a group of 16 threads) load data from the same memory
segment in a single transaction: this is memory coalescing. It can be
seen that the array-of-structures ordering used for the distribution
in the CPU code would not be suitable for coalescing; in fact, it would
result in serialised memory accesses and relative poor performance.
To meet the coalescing criteria and allow consecutive threads to read
consecutive memory addresses on the GPU, we transpose the layout of the
distribution so that, for each velocity component, consecutive sites
are contiguous in memory (``structure-of-arrays'' order). A schematic of
the GPU collision code is given in Listing~\ref{ch14:listing2}.

\begin{lstlisting}[float, label=ch14:listing2,
caption = Collision schematic for GPU.]
/* compute current site index 'is' from CUDA thread and block */
/* indices and load distribution from "structure-of-arrays" */

for (i = 0; i < 19; i++)    
  f[i] = f_soa[nsites*i + is]

/* perform matrix-vector multiplication as before */
...
\end{lstlisting}

\textit{Ludwig} was modified to allow a choice of distribution data layout
at compilation time depending on the target architecture: CPU or GPU. We
defer some further comments on software engineering aspects of the code to
the summary.

\section{Multiple GPU Implementation}\label{ch14:sec:parallelgpu}

To execute on multiple GPUs, we use the same domain decomposition
and message passing framework as the CPU version. Within each
sub-domain (allocated to one MPI task) the GPU implementation
proceeds as described in the previous section. The only additional
complication is that halo transfers between GPUs must be staged
through the host (in future, direct GPU to GPU data transfers via
MPI may be possible, obviating the need for these steps). This
means host MPI sends must be preceded by appropriate device to host
transfers and host MPI receives must be followed by corresponding host
to device transfers.

%To execute on multiple GPUs in parallel we decompose the problem using
%the existing high-level parallelisaton framework: each MPI task acts
%as a host to a separate GPU, with each GPU kernel operating on the
%subset of the lattice volume corresponding to the partition assigned
%to that task.  As described in Section \ref{ch14:sec:background}, halo
%data exchanges are required at each timestep.  Significant
%developments are made for the GPU adaptation of this communication
%phase to achieve good inter-GPU communication performance and allow
%scaling to many GPUs.

%The CPU code uses strided MPI datatypes to define the six
%planes of lattice data to be sent or received, and performs the
%communications {\it in-place}, i.e.  there is no explicit buffering of
%data in the application.  For the GPU adaptation, the halo planes must
%be exchanged between GPUs, on which MPI functionality is not
%available, so transfers must be staged through the host CPUs. Buffer
%packing/unpacking of the halo planes is explicitly performed using
%CUDA kernels, and CUDA memory copy calls are used to transfer data
%from a device to its host (using pinned host memory) through the PCI-e
%bus (and vice versa). MPI calls are used to transfer the buffers
%between hosts. The CPU version of Ludwig has a mechanism, using MPI built-in
%data types, to reduce the amount of data communicated by around a
%factor of four by only sending those elements of the distribution
%propagating outward from a local domain. For the GPU adaptation, this
%filtering of data is explicitly performed (again since MPI
%functionality is not available on the GPU) in the buffer
%packing/unpacking kernels.  Special care had to be taken for those
%lattice sites on the corners of the halo planes, which had to be sent
%in more than one direction (since the propagation includes accesses to
%diagonal neighbours).

In practice, this data movement requires additional GPU kernels to
pack and unpack the relevant data before and after corresponding MPI
calls. However, the standard shift algorithm, in which each co-ordinate
direction is treated in turn, does provide some scope for the
overlapping of different operations.
For example, after the data for the first co-ordinate direction have been
retrieved by the host, these can be exchanged using MPI between
hosts at the same time as kernels for packing and retrieving of
data for the second co-ordinate direction are
executed. This
overlapping must respect the synchronisation required to ensure
that data values at the corners of the sub-domain are transferred
correctly.
We use a separate CUDA stream for each co-ordinate direction:
this allows some of the host-device communication time to be
effectively ``hidden'' behind the host-host MPI communication,
resulting in an overall speedup.
The improvement is more pronounced for the smaller local lattice size,
perhaps because of less CPU memory bandwidth contention. The
overlapping is then a particularly valuable aid to strong scaling,
where the local system size decreases as the number of GPUs increases.

%\section{Performance}\label{ch14:sec:performance}

To demonstrate the effectiveness of this approach, we compare the
performance of both CPU and GPU versions of \textit{Ludwig}. To
test the complex fluid nature of the code, the problem is
actually an immiscible fluid mixture which uses a second distribution
function to introduce a composition variable. The interested reader
is referred to \cite{ch14:stratford-jsp2005} for further details. The
largest total problem size used is $1548\times 1764\times 1568$ ?PLEASE CHECK EXACTLY.
The CPU system features a Cray XE6 architecture with 2 16-core AMD Opteron
CPUs per node, and with nodes interconnected using Cray Gemini technology.
For GPU results, a Cray XK6 system is used: this is very similar to the
XE6, but has one CPU per node replaced with an NVIDIA X2090 GPU. Each
node in the GPU system therefore features a single Opteron CPU acting
as a host to a single GPU. The inter-node interconnect architecture is
the same as for the Cray XE6. The GPU performance tests use a prototype
Cray XK6 system with 936 nodes (the LARGEST AVAILABLE AT TIME OF WRITING?).
To provide a fair comparison, we compare scaling on a \textit{per node}
basis. That is, we compare 1 fully occupied 32-core CPU node
(running 32 MPI tasks) with 1 GPU node (host running 1 MPI task,
and 1 device). We believe this is representative of the true ``cost'' of
a simulation in terms of accounting budgets, or electricity.
